A comparison of model-based and design-based approaches for spatial data.

### 2021-05-20 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
updates to "spatially balanced design and analysis"

---
title: Short Paper
author:
  - name: Alice Anonymous
    email: alice@example.com
    affiliation: Some Institute of Technology
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
abstract: |
  This is the abstract.

@@

journal: "An awesome journal"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
#linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
output: rticles::elsevier_article
---

@@

Introduction
==========

@@

Design-Based Overview \citep{Sarndal2003model, Lohr2009sampling}

@@

Model-Based Overview \citep{Cressie2015statistics, Schabenberger2017statistical}

@@

Design-Based and Model-Based Comparisons \citep{Hansen1983evaluation, Brus1997random, VerHoef2002sampling, Cooper2006sampling, sterba2009alternative, Brus2020statistical, Chan2020bayesian}

@@

\citep{StevensOlsen2003VarianceEstimation, StevensOlsen2004GRTS}

@@

\citep{VerHoef2002sampling, VerHoef2008spatial, Higham2020adjusting}

Finite Population Block Kriging (FPBK) is an alternative to samipling-based methods [@VerHoef2008spatial]. FPBK expands the geostatistical kriging framework to the finite population setting. Instead of basing inference off of a specific sampling design, we assume the data were generated by a spatial process with parameters that can be estimated using the framework of a model.

@VerHoef2008spatial gives details on the theory of FPBK, but some of the basic principles are summarized below. For a response variable $\mathbf{z}$ that can be measured on a finite number of $N$ sites, we want to predict some linear function of the response variable, $\tau(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}$ is a vector of weights. For example, if we want to predict the total abundance across all sites, then we would use a vector of 1's for the weights.

@@

We also typically assume that there is spatial correlation in $\bm{\delta}$, which can be modeled using a covariance function. Many common choices for this function assume that spatial covariance decreases with increasing Euclidean distance between sites. The primary function used throughout the simulations and applications of this manuscript is the Exponential covariance function: the $i,j^{th}$ entry for $\var(\bm{\delta})$ is
\mbox{}
\begin{equation}
\cov(\delta_i, \delta_j) = \theta_3 + \theta_1\exp(-h_{i,j}/\theta_2), \label{equation:expcov}
\end{equation}
where $h_{i,j}$ is the distance between sites $i$ and $j$, and $\bm{\theta}$ is a vector of spatial covariance parameters of the partial sill $\theta_1$, the range $\theta_2$, and the nugget $\theta_3$. However, any spatial covariance function could be used in the place of the Exponential, including functions that allow for anisotropy [pg. 80 - 93][@ChilesEtAl1999GeostatisticsModelingSpatial].

With the above model formulation, the Best Linear Unbiased Predictor (BLUP) for $\tau(\mathbf{b}'\mathbf{z})$ and its prediction variance can be computed. While details of the derivation are in [@VerHoef2008spatial], we note here that the predictor and its variance are both moment-based. Neither require a particular distribution for $\mathbf{z}$.

__Software Implementation__: This probably goes in a different spot (right before simulations?), but putting it here while it's on my mind.

FPBK can be readily performed in `R` with the `sptotal` package [@sptotal2020]. We use `sptotal` for both the simulation analysis and the application, estimating parameters with Restricted Maximum Likelihood (REML).

__Extra Stuff__

References from Brus2020statistical

* de Gruijter, J. J., & ter Braak, C. J. F. (1990). Model-free estimation from spatial samples: A reappraisal of classical sampling theory. Mathematical Geology, 22, 407–415
    * contains info on why classical sampling works with spatially correlated data.

* Brus, D. J., & de Gruijter, J. J. (1997). Random sampling or
geostatistical modelling? Choosing between design-based and
model-based sampling strategies for soil (with discussion). Geoderma, 80, 1–59
    * more comparison between the two appraoches

* I stopped at 2.3

#### Installation

If the document class *elsarticle* is not available on your computer,
you can download and install the system package *texlive-publishers*
(Linux) or install the LaTeX package *elsarticle* using the package
manager of your TeX installation, which is typically TeX Live or MikTeX.

#### Usage

Once the package is properly installed, you can use the document class
*elsarticle* to create a manuscript. Please make sure that your
manuscript follows the guidelines in the Guide for Authors of the
relevant journal. It is not necessary to typeset your manuscript in
exactly the same way as an article, unless you are submitting to a
camera-ready copy (CRC) journal.

#### Functionality

The Elsevier article class is based on the standard article class and
supports almost all of the functionality of that class. In addition, it
features commands and options to format the

-   document style

-   baselineskip

-   front matter

-   keywords and MSC codes

-   theorems, definitions and proofs

-   lables of enumerations

-   citation style and labeling.

Front matter
============

The author names and affiliations could be formatted in two ways:

(1) Group the authors per affiliation.

(2) Use footnotes to indicate the affiliations.

See the front matter of this document for examples. You are recommended
to conform your choice to the journal you are submitting to.

Bibliography styles
===================

There are various bibliography styles available. You can select the
style of your choice in the preamble of this document. These styles are
Elsevier styles based on standard styles like Harvard and Vancouver.
Please use BibTeX to generate your bibliography and include DOIs
whenever available.

Here are two sample references: @Feynman1963118 [@Dirac1953888].

References {#references .unnumbered}
==========

>>>>

---
title: A comparison of model-based and design-based approaches for spatial data.
author:
  - name: Michael Dumelle
    email: Dumelle.Michael@epa.gov
    affiliation: USEPA
    footnote: 1
  - name: Matthew Higham
    email: mhigham@stlaw.edu
    affiliation: STLAW
    footnote: 1
  - name: Lisa Madsen
    affiliation: OSU
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Jay M. Ver Hoef
    affiliation: NOAA
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Math, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  This is the abstract.

@@

journal: "An awesome journal"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
#linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
output: rticles::elsevier_article
---

@@

# Introduction

Please leave comments in your \textcolor{blue}{c}\textcolor{magenta}{o}\textcolor{green}{l}\textcolor{red}{o}\textcolor{cyan}{r}: \textcolor{blue}{Michael}, \textcolor{magenta}{Matt}, \textcolor{green}{Lisa}, \textcolor{red}{Tony}, \textcolor{cyan}{Jay}.

# Background

@@

Design-Based Overview

<!-- \citep{Sarndal2003model, Lohr2009sampling} -->

@@

Model-Based Overview

<!-- \citep{Cressie2015statistics, Schabenberger2017statistical} -->

@@

Design-Based and Model-Based Comparisons

<!-- \citep{Hansen1983evaluation, Brus1997random, VerHoef2002sampling, Cooper2006sampling, sterba2009alternative, Brus2020statistical, Chan2020bayesian} -->

@@

Spatially balanced sampling algorithms use spatial information to obtain samples spread out in space. Spatially balanced samples are useful because they tend to yield estimators that are more precise than estimators constructed from an sampling algorithm that is not spatially balanced ([@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @benedetti2017spatiallyreview; @wang2013design]). Many spatially balanced sampling algorithms exist, including the Generalized Random Tessellation Stratified [@stevens2004spatially], the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance [@benedetti2017spatially], and Halton Iterative Partitioning [@robertson2018halton] algorithms. Here we focus on the Generalized Random Tessellation Stratified (GRTS) algorithm, which has several attractive properties that we discuss next.

The GRTS algorithm is used to sample from finite and infinite sample frames. A finite sample frames contains a finite number of sampling units and is related to a point geometry. An infinite sample frame contains an infinite number of sampling units and is related to linear and polygon geometries. Examples of point, linear, and polygon resources include lake centroids, stream networks, and wetland areas, respectively. In addition to its applicability for finite and infinite sample frames, the GRTS algorithm naturally accommodates stratified designs and designs with unequal selection probabilities. The algorithm has also been used to select replacement sites using reverse hierarchical ordering [@stevens2004spatially]. Replacement sites are used to replace sites in the original sample that cannot be sampled, often as a result of physical difficulty in reaching the site or landowner denial of access to the sites. More recently, the GRTS algorithm also accommodates legacy (historical) sites, minimum distance between sites, and nearest neighbor replacement sites. The GRTS algorithm is implemented in the \textbf{\textsf{R}} package \texttt{spsurvey} [@dumelle2021spsurvey].

describe grts

discuss spatial balance

describe estimation and local neighborhood variance

<!-- \citep{StevensOlsen2003VarianceEstimation, StevensOlsen2004GRTS} -->

@@

<!-- \citep{VerHoef2002sampling, VerHoef2008spatial, Higham2020adjusting} -->

Finite Population Block Kriging (FPBK) is an alternative to samipling-based methods [@verhoef2008spatial]. FPBK expands the geostatistical kriging framework to the finite population setting. Instead of basing inference off of a specific sampling design, we assume the data were generated by a spatial process with parameters that can be estimated using the framework of a model.

@verhoef2008spatial gives details on the theory of FPBK, but some of the basic principles are summarized below. For a response variable $\mathbf{z}$ that can be measured on a finite number of $N$ sites, we want to predict some linear function of the response variable, $\tau(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}$ is a vector of weights. For example, if we want to predict the total abundance across all sites, then we would use a vector of 1's for the weights.

@@

We also typically assume that there is spatial correlation in $\bm{\delta}$, which can be modeled using a covariance function. Many common choices for this function assume that spatial covariance decreases with increasing Euclidean distance between sites. The primary function used throughout the simulations and applications of this manuscript is the Exponential covariance function: the $i,j^{th}$ entry for $\var(\bm{\delta})$ is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = \theta_3 + \theta_1\exp(-h_{i,j}/\theta_2), \\
\textcolor{blue}{\sigma^2[(1 - v)\exp(-3\mathbf{h}_{i,j}/\phi) + v\mathbbm{1}\{\mathbf{h}_{i,j} = 0\}]}, 
\end{align}
where $h_{i,j}$ is the distance between sites $i$ and $j$, and $\bm{\theta}$ is a vector of spatial covariance parameters of the partial sill $\theta_1$, the range $\theta_2$, and the nugget $\theta_3$. However, any spatial covariance function could be used in the place of the Exponential, including functions that allow for anisotropy [pg. 80 - 93][@chiles1999geostatistics].

With the above model formulation, the Best Linear Unbiased Predictor (BLUP) for $\tau(\mathbf{b}'\mathbf{z})$ \textcolor{blue}{Did you mean to give the form of the BLUP here? $\tau(\mathbf{b}'\mathbf{z})$ is vague} and its prediction variance can be computed. While details of the derivation are in [@verhoef2008spatial], we note here that the predictor and its variance are both moment-based. Neither require a particular distribution for $\mathbf{z}$.

# Numerical Study

## Software

FPBK can be readily performed in `R` with the `sptotal` package [@higham2020sptotal]. We use `sptotal` for both the simulation analysis and the application, estimating parameters with Restricted Maximum Likelihood (REML).

# Discussion

<!-- References from Brus2020statistical, de1990model, Brus1997random -->

# References {#references .unnumbered}
